---
title: "Phenology sensitivity based on PhenoCam sites in CONUS"
output: html_notebook
---

### Site map & PFT distribution

```{r echo=TRUE}
library(tidyverse)
library(arrow)
library(rnaturalearth)
library(sf)
theme_set(theme_minimal())

site_meta <- read_parquet("phenocam_site_in_usa.parquet")

usa <- ne_countries(scale = "medium", country = "united states of america", returnclass = "sf")

site_points <- st_as_sf(site_meta, coords = c("lon", "lat"), crs = 4326)
bbox <- st_bbox(site_points)

ggplot() +
  geom_sf(data = usa, fill = "#f2f2f2", color = "black") +  # Base map
  geom_sf(data = site_points, mapping=aes(color=primary_veg_type)) +    # Your points
  coord_sf(xlim = c(bbox[1], bbox[3]), ylim = c(bbox[2], bbox[4])) +             # Zooming box for US mainland
  theme_minimal() +
  labs(color="Primary PFT")
```

```{r}
site_meta %>%
  group_by(primary_veg_type) %>%
  summarize(site_years = sum(site_years))
```

### Summarize annual phenology

Here we are modeling the annual phenology cycle as a double logistic. Greenup and senescence are both controlled by an onset date and a slope parameter that describes the speed of the transition. The form of the logistic equation for each transition is:

$$
f(x) = \frac{L}{1+\exp(-k(x-\beta))}
$$ 

where $k$ is the slope parameter, $L$ is the upper asymptote of the function, and $\beta$ is the midpoint of the transition. Note that $f(\beta)$ is half of its maximum value. $k > 0$ implies the transition is positive (greenup), while $k < 0$ implies the transition is negative (senescence). Since it is not clear how well green chromatic coordinate, the greenness metric used by PhenoCam sites, translates to vegetation activity, we will assume that $L=1$.

Using the subscripts `up` and `dn` to describe greenup and senescence, respectively, the full phenological cycle is given by:

$$
f(x) = \frac{1}{1+\exp(-k_{up}(x-\beta_{up}))} \times 1 - \frac{1}{1+\exp(-k_{dn}(x-\beta_{dn}))}
$$
The below plot shows a curve where $k_{up} = -k_{dn} = -0.1$, $\beta_{up} = 25$, and $\beta_{dn} = 75$.

```{r}
double_logistic <- function(x, kup, kdn, betaup, betadn) {
  log1 = 1 / (1 + exp(-kup * (x - betaup)))
  log2 = 1 / (1 + exp(-kdn * (x - betadn)))
  
  return(log1 * (1-log2))
}

kup = 0.2
kdn = 0.3
betaup = 25
betadn = 75

x <- seq(0, 100)

ggplot(NULL) +
  geom_line(mapping=aes(x=x, y=double_logistic(x, kup, kdn, betaup, betadn))) +
  theme_minimal()
```

PhenoCam sites give us the date of transition and relative greenness at 10%, 25%, and 50% of the way through greenup and senescence. Using the 10% and 50% points, the sites give us $x_{10}$, $x_{50}$, $f(x_{10})$, and $f(x_{50})$. We note that $x_{50} = x_{mid}$ and can solve for $k$ within each transition:

$$
k = \ln \left( \frac {2f(x_{50})} {f(x_{10})} - 1 \right) \times \frac{1}{x_{50} - x_{10}}
$$
To ensure parameters are on the same scale between sites, $f(x)$ must be relativized to the full greenness range within the site.

```{r summarize_phenology}

site_pheno <- read_parquet("phenocam_site_in_usa_phenology.parquet")
gcc_for_analysis <- "gcc_mean"

site_pheno_clean <- site_pheno %>%
  filter(gcc_value == gcc_for_analysis) %>%
  mutate(rel_green_50 = (threshold_50 - min_gcc) / (max_gcc - min_gcc),
         rel_green_10 = (threshold_10 - min_gcc) / (max_gcc - min_gcc),
         # Clamp to min/max
         rel_green_50 = pmin(rel_green_50, 1),
         rel_green_10 = pmax(rel_green_10, 0),
         year_of_transition = case_when(
           direction == "falling" ~ pmin(year(parse_date(transition_50)), year(parse_date(transition_10))),
           direction ==  "rising" ~ pmax(year(parse_date(transition_50)), year(parse_date(transition_10)))
        ),
        yday_transition_50 = yday(parse_date(transition_50)),
        yday_transition_10 = yday(parse_date(transition_10))
  ) %>%
  # Sanitize ydays that wrap across years. This occurs when transition_10 >
  # transition_50 in rising transition, or when transition_50 > transition_10
  # in falling transition.
  mutate(
    yday_transition_10 = case_when(
      (direction == "falling") & (yday_transition_50 > yday_transition_10) ~ yday_transition_10 + 365,
      .default = yday_transition_10
    ),
    yday_transition_10 = case_when(
      (direction == "rising") & (yday_transition_10 > yday_transition_50) ~ yday_transition_10 - 365,
      .default = yday_transition_10
    )
  ) %>%
  # Calculate slope of each transition
  mutate(
    transition_slope = (
      abs(log((2 * rel_green_50 / rel_green_10) - 1) / 
            (yday_transition_50 - yday_transition_10))
    )
  )

```

### Summarize annual weather

Although phenology is most sensitive to the weather immediately preceding the transition, we are interested in how phenology timing varies at continental scale. For this reason, we will derive phenology transitions from annual weather summaries.

```{r}
site_climate <- read_parquet("phenocam_site_in_usa_climate.parquet")

site_climate_jfm <- site_climate %>%
  filter(yday < 90) %>%
  group_by(year, site) %>%
  summarize(ppt = sum(`prcp (mm/day)`),
            txm = mean(`tmax (deg c)`),
            tnm = mean(`tmin (deg c)`),
            vpm = mean(`vp (Pa)`))

site_climate_ond <- site_climate %>%
  filter(yday > 270) %>%
  group_by(year, site) %>%
  summarize(ppt = sum(`prcp (mm/day)`),
            txm = mean(`tmax (deg c)`),
            tnm = mean(`tmin (deg c)`),
            vpm = mean(`vp (Pa)`))

site_climate_pheno <- site_pheno_clean %>%
  left_join(site_climate_jfm,
            by=c(
              "site"="site",
              "year_of_transition"="year"
            )) %>%
  left_join(site_meta,
            by=c("site"="sitename"))
```

### Mixed effects model

```{r}
library(lme4)

site_climate_pheno %>%
  filter(direction == "rising") %>%
  ggplot(aes(x=txm, y=yday_transition_50)) +
  geom_point(mapping=aes(color=primary_veg_type), alpha=0.25) +
  geom_smooth(mapping=aes(color=primary_veg_type, group=primary_veg_type), method="lm") +
  labs(
    x=expression("JFM mean Tx ("*degree*"C)"), 
    y="50% rising greenness transition, day of year",
    color=""
  ) +
  theme_bw()
```